---
title: "Conserve the Bees"
author: "F.J. Morales"
doi: "GBIF.org (05 January 2024) GBIF Occurrence Download  https://doi.org/10.15468/dl.kktnnk"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

Install packages
```{r install packages, eval=FALSE, include=FALSE}
install.packages("tidyverse") #cleaning, exploring, analyzing and visualizing data
install.packages("rgbif") #communicating with GBIF API
install.packages("DBI") #connecting to a database server
install.packages("RPostgres") #communicating with PostgreSQL
install.packages("maps") #converting coordinates to states
install.packages("taxize") #parsing scientific names
```

Load libraries
```{r load libraries}
library(tidyverse)
library(rgbif)
library(DBI)
library(RPostgres)
library(maps)
library(taxize)
```

Store bee family names on a vector, get their **backbone** and **taxonKey** data
```{r family}
familyName <-
  c(
    'Apidae',
    'Halictidae',
    'Megachilidae',
    'Andrenidae',
    'Colletidae',
    'Melittidae',
    'Stenotritidae'
  )
familyBackbone <- name_backbone_checklist(familyName)
familyKey <- familyBackbone$usageKey
```

Get download from **GBIF** query and import it into ***RStudio***
```{r gbif, eval=FALSE, include=FALSE}
gbifGet <-
  occ_download_get("0056337-231120084113126",
                   path = ".",
                   overwrite = FALSE)
gbifImport <- occ_download_import(gbifGet)
```

Connect with **PostgreSQL** database
```{r postgres, eval=FALSE, include=FALSE}
postgres <- dbConnect(
  RPostgres::Postgres(),
  dbname = 'beeDB',
  host = 'localhost',
  port = 5432,
  user = rstudioapi::askForPassword("Please enter user:"),
  password = rstudioapi::askForPassword("Please enter password:")
)
```

Store occurrences dataset on database
```{r create table, eval=FALSE, include=FALSE}
dbWriteTable(postgres, "occurrences", gbifImport)
```

Rename columns
```{sql rename columns, eval=FALSE, include=FALSE}
ALTER TABLE occurrences RENAME COLUMN "gbifID" TO gbifid_;
ALTER TABLE occurrences RENAME COLUMN family TO family_;
ALTER TABLE occurrences RENAME COLUMN "scientificName" TO scientificname_;
ALTER TABLE occurrences RENAME COLUMN "occurrenceStatus" TO status_;
ALTER TABLE occurrences RENAME COLUMN "individualCount" TO count_;
ALTER TABLE occurrences RENAME COLUMN "decimalLatitude" TO latitude_;
ALTER TABLE occurrences RENAME COLUMN "decimalLongitude" TO longitude_;
ALTER TABLE occurrences RENAME COLUMN "eventDate" TO date_;
```

Query the dataset into a **dataframe**
```{r query, eval=FALSE, include=FALSE}
# connection=postgres
occurrencesDataframe <- dbGetQuery(
  postgres,
  "SELECT
    gbifid_,
    family_,
    scientificname_,
    count_,
    latitude_,
    longitude_,
    date_
  FROM
    occurrences
  WHERE
    status_ = 'PRESENT' AND
    date_ >= '1923-01-01 00:00:00' AND
    date_ <= '2023-12-31 23:59:59'
  ORDER BY
    date_ ASC"
) %>%
  mutate(date_ = as_date(date_))

str(occurrencesDataframe)
```

```{r read dataset}
occurrencesDataframe <- read_csv("occurrencesDataframe.csv") %>%
  mutate(date_ = as_date(date_))

occurrencesDataframe$...1 <- NULL
str(occurrencesDataframe)
```

_Mutate_ **count_** field to replace nulls with 1 and _floor_ **date_** to month
```{r count and date}
occurrencesDataframe <- occurrencesDataframe %>%
  mutate(count_ = replace_na(count_, 1),)

occurrencesDataframe$date_ <- occurrencesDataframe$date_ %>%
  floor_date(unit = "month")
```

Get **state_** and **county_** from coordinates (**latitude_** and **longitude_**)
```{r coordinates}
occurrencesDataframe <- occurrencesDataframe[!is.na(occurrencesDataframe$latitude_ | occurrencesDataframe$longitude_),]
occurrencesDataframe$locality_ <- map.where(database  = "county", occurrencesDataframe$longitude_, occurrencesDataframe$latitude_)
occurrencesDataframe <- occurrencesDataframe %>%
  relocate(locality_, .after = count_) %>%
  separate_wider_delim(locality_, delim = ",", names = c("state_", "county_"))

occurrencesDataframe$state_ <- str_to_title(occurrencesDataframe$state_)
occurrencesDataframe$county_ <- str_to_title(occurrencesDataframe$county_)
```

Establish occurrences location
```{r location}
#location <- str_to_title(readline(prompt = "Please enter location: "))
location <- c('North Carolina')
occurrencesTibble <- occurrencesDataframe %>%
  filter(occurrencesDataframe$state_ == location)

tibble(occurrencesTibble)
```

Get **genus_** and **species_** (genus and epithet) from **scientificname_**
```{r scientific names}
names <- tibble(name_parse(scientificname = occurrencesTibble$scientificname_, curlopts = list()))
names <- tibble(species_ = names$canonicalname,
                genus_ = names$genusorabove,
                epithet_ = names$specificepithet)

names <- replace_na(names, list(genus_ = "Gen.", epithet_ = "sp."))

occurrencesTibble$genus_ <- names$genus_
occurrencesTibble$species_ <- names$epithet_

occurrencesTibble$species_[occurrencesTibble$species_ != "sp."] <- 
  paste(str_sub(occurrencesTibble$genus_[occurrencesTibble$species_ != "sp."], start = 1, end = 1),
        occurrencesTibble$species_[occurrencesTibble$species_ != "sp."],
        sep = ". ")

occurrencesTibble$species_[occurrencesTibble$species_ == "sp."] <- 
  paste(occurrencesTibble$genus_[occurrencesTibble$species_ == "sp."],
        occurrencesTibble$species_[occurrencesTibble$species_ == "sp."], sep = " ")

occurrencesTibble <- occurrencesTibble %>%
  relocate(genus_, .after = family_)

occurrencesTibble <- occurrencesTibble %>%
  relocate(species_, .after = genus_)
```

```{r summarize time}
summaryTime <- aggregate(occurrencesTibble$count_, list(occurrencesTibble$date_), FUN = sum) %>%
  rename(date_ = Group.1, sum_ = x)
```

```{r summarize space}
summarySpace <- aggregate(occurrencesTibble$count_, list(occurrencesTibble$county_), FUN = sum) %>%
  rename(county_ = Group.1, sum_ = x)
```

